Notes on Validated Model Counting \\ Version of \today

### 2022-04-08 doc/mc-notes.tex
New proof framework

\newcommand{\one}{\mbox{\bf 1}}
\newcommand{\zero}{\mbox{\bf 0}}
\newcommand{\leafone}{L_1}
\newcommand{\leafzero}{L_0}
\newcommand{\booland}{\land}
\newcommand{\boolor}{\lor}
\newcommand{\boolxor}{\oplus}
\newcommand{\boolnot}{\neg}
%\newcommand{\tautology}{\top}
%\newcommand{\nil}{\bottom}
\newcommand{\tautology}{1}
\newcommand{\nil}{0}
\newcommand{\obar}[1]{\overline{#1}}
\newcommand{\ite}{\mbox{\it ITE}}

@@

From
this we can define the {\em negation} of function $f$ as the function
$\boolnot f$ such that
$\interpset{\boolnot f} = \{ \interp | f(\interp) = 0\}$.
We can also define the conjunction and disjunction operations over functions $f_1$ and $f_2$ as characterized by the sets
$\interpset{f_1 \booland f_2} = \interpset{f_1} \cap \interpset{f_2}$ and
$\interpset{f_1 \boolor f_2} = \interpset{f_1} \cup \interpset{f_2}$.

@@

Let $\ring$ denote the elements of a commutative ring.  A {\em
  separable cost function} $\cost:X \rightarrow \ring$ assigns a value
from the ring to each variable.  We extend this function by defining
the cost of literal as $\cost(\obar{x}_i) = 1 - \cost(x_i)$, the cost
of an assignment as $\cost(\interp) = \prod_{l_i \in \interp}
\cost(l_i)$, and the cost of a function $f$ as $\cost(f) =
\sum_{\interp \in \interpset{f}} \cost(\interp)$.

{\bf Example 1}: Let $\ring$ be the set of rational numbers and
$\cost(x_i) = 1/2$ for all variables $x_i$.  Then cost of every
assignment will be $1/2^{n}$, and the cost of a function will be its
{\em density}, denoted $\density(f)$.  That is $\density(f) \in [0,
  1]$ will be the fraction of assignments for which the function
evaluates to 1.  Given an ability to compute the density of
function $f$, we can scale this by $2^n$ to compute $|\interpset{f}|$, the core task of model counting.
Using density as our metric has the advantage that it
does not vary when the function is embedded in a larger domain $X'
\supseteq X$.

{\bf Example 2}: Let $\ring$ be a field with $|\ring| > 2n$,
and let $\hashset$ be the set functions
mapping elements of $X$ to elements of $\ring$.  For
two distinct functions $f_1$ and $f_2$ and a randomly chosen $\fhash
\in \hashset$, the probability that $h(f_1) = h(f_2)$ will be at most
$2^n/|\ring|< 1/2$.  Therefore, these functions can be used as part of a
randomized algorithm for equivalence testing~\cite{blum:ipl:1980}.

@@

\begin{lemma}[Complementation]
  For separable cost function $\cost$ and Boolean function $f$:
  $\cost(\neg f) = 1 - \cost(f)$.
\end{lemma}

\begin{lemma}[Conjunction over Independent Domains]
  For separable cost function $\cost$ and Boolean functions $f_1$ and $f_2$ such that $\depend(f_1) \cap \depend(f_2) = \emptyset$:
    $\cost(f_1 \land f_2) = \cost(f_1) \cdot \cost(f_2)$.
\end{lemma}

\begin{lemma}[Disjunction over Disjoint Domains]
  For separable cost function $\cost$ and Boolean functions $f_1$ and $f_2$ such that $\interpset{f_1} \cap \interpset{f_2} = \emptyset$:
    $\cost(f_1 \lor f_2) = \cost(f_1) + \cost(f_2)$.
\end{lemma}

@@

\begin{lemma}
  \label{lemma:negation}
  For any formula $F$:
  \begin{eqnarray*}
    \mcount(\neg F) & =  & 2^n - \mcount(F)
  \end{eqnarray*}
\end{lemma}

@@

\begin{theorem}
\label{theorem:count}
Letting $n = |X|$, the following recursive formula holds for any node in a VPITE graph:
\begin{eqnarray}
\mcount(\tautology) & = & 2^{n} \label{eqn:count:true}\\
\mcount(\nil) & = & 0 \label{eqn:count:false}\\
\mcount(x) & = & 2^{n-1} \label{eqn:count:variable}\\
\mcount[\ite(v_i, v_t, v_e)] & = & \frac{\mcount(v_i) \cdot \mcount(v_t) + [2^n-\mcount(v_i)] \cdot \mcount(v_e)}{2^n} \label{eqn:count:ite}
\end{eqnarray}  
\end{theorem}
\noindent
{\em Proof:} The key result here is (\ref{eqn:count:ite}).  It follows
from Lemmas~\ref{lemma:negation}--\ref{lemma:sum}, along with the definition of the ITE operator.

>>>>

\newcommand{\one}{\mbox{\bf 1}}
\newcommand{\zero}{\mbox{\bf 0}}
\newcommand{\leafone}{L_1}
\newcommand{\leafzero}{L_0}
\newcommand{\booland}{\land}
\newcommand{\boolor}{\lor}
\newcommand{\pand}{\mathbin{\land_{\rm pv}}}
\newcommand{\por}{\mathbin{\lor_{\rm pa}}}
\newcommand{\boolxor}{\oplus}
\newcommand{\boolnot}{\neg}
%\newcommand{\tautology}{\top}
%\newcommand{\nil}{\bottom}
\newcommand{\tautology}{1}
\newcommand{\nil}{0}
\newcommand{\obar}[1]{\overline{#1}}
\newcommand{\ite}{\mbox{\it ITE}}

@@

From
this we can define the {\em complement} of function $f$ as the function
$\boolnot f$ such that
$\interpset{\boolnot f} = \{ \interp | f(\interp) = 0\}$.
We can also define the conjunction and disjunction operations over functions $f_1$ and $f_2$ as characterized by the sets
$\interpset{f_1 \booland f_2} = \interpset{f_1} \cap \interpset{f_2}$ and
$\interpset{f_1 \boolor f_2} = \interpset{f_1} \cup \interpset{f_2}$.

@@

Let $\ring$ denote the elements of a commutative ring.  A {\em
  separable cost function} $\cost:X \rightarrow \ring$ assigns a value
from the ring to each variable.  We extend this function by defining
the cost of literal $\obar{x}_i$ as $\cost(\obar{x}_i) = 1 - \cost(x_i)$, the cost
of an assignment as $\cost(\interp) = \prod_{l_i \in \interp}
\cost(l_i)$, and the cost of a function $f$ as $\cost(f) =
\sum_{\interp \in \interpset{f}} \cost(\interp)$.

{\bf Example 1}: Let $\ring$ be the set of rational numbers and
$\cost(x_i) = 1/2$ for all variables $x_i$.  The cost of every
assignment will be $1/2^{n}$, and the cost of a function will be its
{\em density}, denoted $\density(f)$.  That is $\density(f) \in [0,
  1]$ will be the fraction of assignments for which the function
evaluates to 1.  Given an ability to compute the density of
function $f$, we can scale this by $2^n$ to compute $|\interpset{f}|$, the core task of model counting.
Using density as our metric has the advantage that it
does not vary when the function is embedded in a larger domain $X'
\supseteq X$.  As a variant, we can support some forms of
weighted model counting by assigning weights of the form
$\cost(x_i) = w_i$, where $0 \leq w_i \leq 1$ for every variable $x_i$.

{\bf Example 2}: Let $\ring$ be a field with $|\ring| > 2n$,
and let $\hashset$ be the set of functions
mapping elements of $X$ to elements of $\ring$.  For
two distinct functions $f_1$ and $f_2$ and a randomly chosen $\fhash
\in \hashset$, the probability that $h(f_1) = h(f_2)$ will be at most
$2^n/|\ring|< 1/2$.  Therefore, these functions can be used as part of a
randomized algorithm for equivalence testing~\cite{blum:ipl:1980}.

@@

\begin{lemma}[Complementation]
\label{lemma:complementation}
  For separable cost function $\cost$ and Boolean function $f$:
  $\cost(\neg f) = 1 - \cost(f)$.
\end{lemma}

\begin{lemma}[Variable-Partitioned Conjunction]
\label{lemma:conjunction}
  For separable cost function $\cost$ and Boolean functions $f_1$ and $f_2$ such that $\depend(f_1) \cap \depend(f_2) = \emptyset$:
    $\cost(f_1 \land f_2) = \cost(f_1) \cdot \cost(f_2)$.
\end{lemma}
We use the notation $f_1 \pand f_2$ to denote the conjunction of $f_1$ and
$f_2$ under the condition that $f_1$ and $f_2$ are defined over
disjoint sets of variables.

\begin{lemma}[Assignment-Partitioned Disjunction]
\label{lemma:disjunction}
  For separable cost function $\cost$ and Boolean functions $f_1$ and $f_2$ such that $\interpset{f_1} \cap \interpset{f_2} = \emptyset$:
    $\cost(f_1 \lor f_2) = \cost(f_1) + \cost(f_2)$.
\end{lemma}
We use the notation $f_1 \por f_2$ to denote the disjunction of $f_1$ and $f_2$ under the
condition that $f_1$ and $f_2$ hold for mutually exclusive assignments.

\section{Proof Framework for Cost Functions}

We introduce the CRAT clausal proof framework for generating a
checkable proof of the valuation of a Boolean formula, given in
conjunctive normal form, according to a specified cost function.  A
CRAT proof provides two capabilities: 1) a schema for computing a cost
function using the ring operations, and 2) a proof that this schema
accurately characterizes the cost function for the input formula.
A proof can be either {\em function independent}, meaning that it holds
for any cost function, or {\em function dependent}, meaning that
it holds for only for a specific cost function.

The CRAT format draws its inspiration from the QRAT format for
quantified Boolean formulas (QBF).  The following are its key properties:
\begin{itemize}
\item Clause addition is allowed, as long as either 1) the new clause is
  blocked, or 2) it satisfies the RAT property with respect to a sequence
  of earlier antecedent clauses.
\item  Extension variables can be introduced only according to the operations $\pand$, $\por$, and {\em abstraction}
\begin{itemize}
\item The checker tracks the dependency set for every input and
  extension variable.  When an extension variable is introduced based
  on the $\pand$ operation, the dependency sets of its arguments must
  be disjoint.
\item When an extension variable is introduced based on the $\por$
  operation, the proof step must cite earlier steps providing a RUP proof
  that the two arguments are mutually exclusive.
\item The {\em abstraction} operation allows an extension
  variable to serve as an alias for one of more other 
  variables.  Use of this rule renders the proof function dependent.
\item Boolean complement is provided implicitly by allowing the
  arguments of the extension operations to be literals and not just
  variables.
\end{itemize}
\item Unlike DRAT, the deletion of a clause requires showing that this clause is implied by one or more other clauses.
\item Unlike QRAT, it need not support universal quantification.
\end{itemize}

A CRAT proof follows the same general form as a QRAT satisfaction
proof.  That is, starting with the set of input clauses it produces a
sequence of steps that both add and delete clauses.  At the final
step, there must be only a single unit clause consisting of some
variable or its complement.  Except for trivial cases, the final
variable will be an extension variable.  The sequence of extension
operations define the schema for computing the cost function.  That
is, a value for the cost function is computed for every literal
according to the rules of Lemmas
\ref{lemma:complementation}--\ref{lemma:disjunction}.  For every
abstraction step, all variables having the indicated
extension variable as their alias must evaluate to the same value.

@@

\begin{lemma}
  \label{lemma:complement}
  For any formula $F$:
  \begin{eqnarray*}
    \mcount(\neg F) & =  & 2^n - \mcount(F)
  \end{eqnarray*}
\end{lemma}

@@

\begin{theorem}
\label{theorem:count}
Letting $n = |X|$, the following recursive formula holds for any node in a VPITE graph:
\begin{eqnarray}
\mcount(\tautology) & = & 2^{n} \label{eqn:count:true}\\
\mcount(\nil) & = & 0 \label{eqn:count:false}\\
\mcount(x) & = & 2^{n-1} \label{eqn:count:variable}\\
\mcount[\ite(v_i, v_t, v_e)] & = & \frac{\mcount(v_i) \cdot \mcount(v_t) + [2^n-\mcount(v_i)] \cdot \mcount(v_e)}{2^n} \label{eqn:count:ite}
\end{eqnarray}  
\end{theorem}
\noindent
{\em Proof:} The key result here is (\ref{eqn:count:ite}).  It follows
from Lemmas~\ref{lemma:complement}--\ref{lemma:sum}, along with the definition of the ITE operator.
